---
title: "Wallace et al Psychopharm spiny mouse - October 2023"
author: "Wallace"
date: "10/19/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---



#get packages
```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(viridis)
library(EnvStats)
library(ggbeeswarm)
library(factoextra)
library(jtools)
library(cowlogdata)
library(rstatix)
library(vcd)
library(effsize)
```

#get dataframes
```{r}
df<-read_csv("PLEASE INSERT PATH TO DATA FILE HERE")


#list of measurements
list<-colnames(df)
list<-list[11:59] #removing categorical variables

```

#boxplots with dotplot overlay, graphs relevant to SPSS results
#this section also averages for litter size and sex ratio (with graph for litter size comparison between treatment)
```{r}

#ssexp stimulus time by sex
  dftemp<-df %>% select(sex, ssexp_stimulus_prop)
  colnames(dftemp)<-c('sex', 'ssexp_stimulus_prop')
  dftemp$sex<-as.factor(dftemp$sex)
p<- ggplot(dftemp, aes(x=sex, y=ssexp_stimulus_prop)) +
   geom_boxplot(lwd=1, fill = c("indianred3", "cornflowerblue"))+
   geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 1)+
   theme_classic()+
  theme(axis.text.y = element_text(size = 14)) + theme(axis.text.x = element_text(size = 14))+
  stat_n_text()
 plot(p)

#ssexp stimulus time by treatment
  dftemp<-df %>% select(treatment, ssexp_stimulus_prop)
  colnames(dftemp)<-c('treatment', 'ssexp_stimulus_prop')
  dftemp$treatment<-as.factor(dftemp$treatment)
p<- ggplot(dftemp, aes(x=treatment, y=ssexp_stimulus_prop)) +
   geom_boxplot(lwd=1, fill = c("aquamarine4", "darkgoldenrod3"))+
   geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 1)+
   theme_classic()+
   theme(axis.text.y = element_text(size = 14)) + theme(axis.text.x = element_text(size = 14))+
  stat_n_text()
 plot(p)


#ls fos by treatment
  dftemp<-df %>% select(treatment, ls_fos)
  colnames(dftemp)<-c('treatment', 'ls_fos')
  dftemp$treatment<-as.factor(dftemp$treatment)
p<- ggplot(dftemp, aes(x=treatment, y=ls_fos)) +
   geom_boxplot(lwd=1, fill = c("aquamarine4", "darkgoldenrod3"))+
   geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 1)+
   theme_classic()+
   theme(axis.text.y = element_text(size = 14)) + theme(axis.text.x = element_text(size = 14))+
  stat_n_text()
 plot(p)
 
 
 #poa perk by treatment
  dftemp<-df %>% select(treatment, poa_perk)
  colnames(dftemp)<-c('treatment', 'poa_perk')
  dftemp$treatment<-as.factor(dftemp$treatment)
p<- ggplot(dftemp, aes(x=treatment, y=poa_perk)) +
   geom_boxplot(lwd=1, fill = c("aquamarine4", "darkgoldenrod3"))+
   geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 1)+
   theme_classic()+
   theme(axis.text.y = element_text(size = 14)) + theme(axis.text.x = element_text(size = 14))+
  stat_n_text()
 plot(p)
 
  #pvn perk by treatment
  dftemp<-df %>% select(treatment, pvn_perk)
  colnames(dftemp)<-c('treatment', 'pvn_perk')
  dftemp$treatment<-as.factor(dftemp$treatment)
p<- ggplot(dftemp, aes(x=treatment, y=pvn_perk)) +
   geom_boxplot(lwd=1, fill = c("aquamarine4", "darkgoldenrod3"))+
   geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 1)+
   theme_classic()+
   theme(axis.text.y = element_text(size = 14)) + theme(axis.text.x = element_text(size = 14))+
  stat_n_text()
 plot(p)
 
 
  #poa perk by sex
  dftemp<-df %>% select(sex, pvn_perk)
  colnames(dftemp)<-c('sex', 'pvn_perk')
  dftemp$sex<-as.factor(dftemp$sex)
p<- ggplot(dftemp, aes(x=sex, y=pvn_perk)) +
   geom_boxplot(lwd=1, fill = c("indianred3", "cornflowerblue"))+
   geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 1)+
   theme_classic()+
   theme(axis.text.y = element_text(size = 14)) + theme(axis.text.x = element_text(size = 14))+
  stat_n_text()
 plot(p)
 
 #mean litter size ALL
 dftemp<-df[!duplicated(df$parents),]
 print(mean(dftemp$littersize))
 print(sd(dftemp$littersize))
 
 #mean litter size SIMPLE
 dftempsimple<-dftemp %>% filter(treatment == "simple")
 print(mean(dftempsimple$littersize))
 print(sd(dftempsimple$littersize))

 
 #mean litter size COMPLEX
 dftempcomplex<-dftemp %>% filter(treatment == "complex")
 print(mean(dftempcomplex$littersize))
 print(sd(dftempcomplex$littersize))

 dftemp<-dftemp %>% select(parents, littersize, littersexratiomf)
 
 preg_length<-c(49, 45, 45, 49, 42, 85, 40, 42, 38, 67, 48, 125, 152, 48)
print(mean(preg_length))
 print(sd(preg_length))
 
 littersexratiolist<-c(0, 1, 0, 1, 1, 0, 2, 0, 1, 0, -1, 1, 1, 0)
 print(mean(littersexratiolist))
 print(sd(littersexratiolist))
 shapiro.test(littersexratiolist) #significant, so use wilcoxon test
 print(wilcox.test(littersexratiolist, mu = 0, alternative = "two.sided"))
 dftemp<-data.frame(littersexratiolist)
 
 wilcoxeffsize<- dftemp %>% wilcox_effsize(littersexratiolist ~ 1, mu = 0)
 print(wilcoxeffsize)

#stacked bar graph of litter size by treatment with chi squared test

 littersizelist<-c(2, 3, 2, 3, 3, 2, 3, 2, 3, 2, 3, 2, 2, 2)
 treatmentlist<- c("complex", "simple", "simple", "complex", "simple", "complex", "simple", "complex", "simple", "simple", "simple", "complex", "complex", "complex")  
 
dftemp<-data.frame(littersizelist, treatmentlist)
dftemp$littersizelist<-as.factor(dftemp$littersizelist)
 
freq <- table(dftemp$treatmentlist, dftemp$littersizelist)

#calculating effect size by hand using spss result:
#φ = √(X2 / n)
sqrt(4.667/14)

ggplot(dftemp, aes(treatmentlist, fill = littersizelist)) + 
    geom_bar() +
    theme_classic()+
    scale_fill_manual(values = c("slategrey","slategray3"))+
    ggtitle("litter size by treatment chi squared")
 
```

#PCA of behavior by treatment and sex
```{r}


#Behavior PCA (not including two timepoint exposures, only behavioral profiling assays) by treatment

templist<- c("ssb_total_aggression_prop", "ssb_total_prosocial_prop","osb_total_aggression_prop", "osb_total_prosocial_prop", "of_center_middle_prop", "nor_lower_middle_prop" )

temp<-df %>% select(name, treatment, all_of(templist))
temp$treatment<-as.factor(temp$treatment)

temp<-na.omit(temp)
temp.noname<-subset(temp,select = -c(name,treatment))
res.pca <- prcomp(temp.noname, scale = TRUE)
fviz_eig(res.pca, addlabels=TRUE) + ggtitle("BEHAVIOR PCA BY TREATMENT")
fviz_pca_var(res.pca,
             col.var = "black",
             #gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)+ ggtitle("BEHAVIOR PCA BY TREATMENT") +
            theme(panel.background=element_blank(),
              panel.border=element_blank(),
              panel.grid.major=element_blank(),
              panel.grid.minor=element_blank(),
              plot.background=element_blank())

 pca <- prcomp(temp.noname, center = TRUE,scale = TRUE)
 summary(pca)
 assaybehavpca<-as.data.frame(pca$x)
 assaybehavpca$name<-temp$name
 assaybehavpca$treatment<-temp$treatment
 pca.loadings <- pca$rotation
 assaybehavpca<-assaybehavpca %>% select(PC1, PC2, name, treatment)

p_value_PC1_treatment<-t.test(PC1~treatment, data = assaybehavpca, na.rm = TRUE)
p_value_PC1_treatment<-as.numeric(p_value_PC1_treatment[3])
assaybehavpca %>% cohens_d(PC1 ~ treatment)
p_value_PC2_treatment<-t.test(PC2~treatment, data = assaybehavpca, na.rm = TRUE)
p_value_PC2_treatment<-as.numeric(p_value_PC2_treatment[3])
assaybehavpca %>% cohens_d(PC2 ~ treatment)


ggplot(assaybehavpca, aes(x=PC1, y=PC2, color = treatment)) +
  geom_point(size=6)+
  scale_color_manual(values = c("aquamarine4", "darkgoldenrod3"))+
  ggtitle(paste("BEHAVIOR PCA t-tests\n PC1 by treatment:", round(p_value_PC1_treatment, 3), 
                "\n PC2 by treatment:", round(p_value_PC2_treatment, 3)))+
  geom_hline(yintercept=0,linetype="longdash", color = "gray90")+
  geom_vline(xintercept=0,linetype="longdash", color = "gray90")+
  theme_classic()




#Behavior PCA by sex

templist<- c("ssb_total_aggression_prop", "ssb_total_prosocial_prop","osb_total_aggression_prop", "osb_total_prosocial_prop", "of_center_middle_prop", "nor_lower_middle_prop" )

temp<-df %>% select(name, sex, all_of(templist))
temp$sex<-as.factor(temp$sex)

temp<-na.omit(temp)
temp.noname<-subset(temp,select = -c(name,sex))
res.pca <- prcomp(temp.noname, scale = TRUE)
fviz_eig(res.pca, addlabels=TRUE) + ggtitle("BEHAVIOR PCA BY sex")
fviz_pca_var(res.pca,
             col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)+ ggtitle("BEHAVIOR PCA BY sex")
 pca <- prcomp(temp.noname, center = TRUE,scale = TRUE)
 summary(pca)
 assaybehavpca<-as.data.frame(pca$x)
 assaybehavpca$name<-temp$name
 assaybehavpca$sex<-temp$sex
 pca.loadings <- pca$rotation
 assaybehavpca<-assaybehavpca %>% select(PC1, PC2, name, sex)

p_value_PC1_sex<-t.test(PC1~sex, data = assaybehavpca, na.rm = TRUE)
p_value_PC1_sex<-as.numeric(p_value_PC1_sex[3])
assaybehavpca %>% cohens_d(PC1 ~ sex)
p_value_PC2_sex<-t.test(PC2~sex, data = assaybehavpca, na.rm = TRUE)
p_value_PC2_sex<-as.numeric(p_value_PC2_sex[3])
assaybehavpca %>% cohens_d(PC2 ~ sex)


ggplot(assaybehavpca, aes(x=PC1, y=PC2, color = sex)) +
  geom_point(size=6)+
  scale_color_manual(values = c("violetred4", "skyblue3"))+
  ggtitle(paste("BEHAVIOR PCA t-tests\n PC1 by sex:", round(p_value_PC1_sex, 3), 
                "\n PC2 by sex:", round(p_value_PC2_sex, 3)))+
  geom_hline(yintercept=0,linetype="longdash", color = "gray90")+
  geom_vline(xintercept=0,linetype="longdash", color = "gray90")+
  theme_classic()

```

#stacked bar graphs of time spent in various areas or doing certain behaviors for each of the four tasks (supp figure)
#also includes OF-NOR repeated measures analysis f time spent in lower middle area
```{r}
#OF to NOR time spent in lower middle

dftemp2<-df %>% select(of_lower_middle_prop, nor_lower_middle_prop, name, treatment)
dftemp2$part<-"of"
dftemp2<-rbind(dftemp2, dftemp2)


dftemp2$part[35:68]<-"znor"
dftemp2$of_lower_middle_prop[35:68]<-dftemp2$nor_lower_middle_prop[35:68]
dftemp2<- dftemp2 %>% select(-nor_lower_middle_prop)
colnames(dftemp2)<-c("lower_middle_prop", "name", "treatment","part")

ggplot(dftemp2, aes(x = part, y = lower_middle_prop, group = name, color = treatment)) +
    scale_color_manual(values = c("aquamarine4", "darkgoldenrod3"))+
    geom_line(color = "gray90", lwd = 0.8) + geom_point(size = 4) + ggtitle("p < 0.001 SPSS")+ theme_classic()

#this is just an average across all individuals because SPSS showed no treatment or sex effects

###########OF

cldata(pathtofile = "PLEASE INSERT PATH TO RAW SCORE SHEETS HERE /of", outputdataname = "dfof", outputzonename = "of_list_of_zones_original", factor = "FALSE")
colnames(dfof) <- paste("of", colnames(dfof), sep = "_")
of_list_of_zones <- paste("of", of_list_of_zones_original, sep = "_")

i<-1
for (i in 1:nrow(dfof)){
  dfof$of_total_wall_prop[i]<-  dfof$of_left_wall_prop[i] + dfof$of_right_wall_prop[i] + dfof$of_upper_wall_prop[i] +    
     dfof$of_lower_wall_prop[i]
  dfof$of_total_arena_prop[i]<-dfof$of_lower_middle_prop[i] + 
     dfof$of_lower_left_prop[i] +               
     dfof$of_upper_middle_prop[i] +              
     dfof$of_upper_right_prop[i] +              
     dfof$of_upper_left_prop[i] +                
     dfof$of_center_left_prop[i] +              
     dfof$of_center_right_prop[i] +
     dfof$of_lower_right_prop[i]}
condition <- rep(c("wall" , "arena" , "center"))
value <- c(mean(dfof$of_total_wall_prop),mean(dfof$of_total_arena_prop),mean(dfof$of_center_middle_prop))
data <- data.frame(condition,value)

ggplot(data, aes(fill=condition, y=value, x = 1)) + 
    geom_bar(position="stack", stat="identity") +
    theme_classic()+
    scale_fill_manual(values = c("orangered4", "indianred3", "rosybrown3"))+
    ggtitle("Open Field Zones 
            (averaged across all)")






############NOV


cldata(pathtofile = "PLEASE INSERT PATH TO RAW SCORE SHEETS HERE /nor", outputdataname = "dfnor", outputzonename = "nor_list_nor_zones_original", factor = "FALSE")
colnames(dfnor) <- paste("nor", colnames(dfnor), sep = "_")
nor_list_nor_zones <- paste("nor", nor_list_nor_zones_original, sep = "_")

i<-1
for (i in 1:nrow(dfnor)){
  dfnor$nor_total_wall_prop[i]<-  dfnor$nor_left_wall_prop[i] + dfnor$nor_right_wall_prop[i] + dfnor$nor_upper_wall_prop[i] +    
     dfnor$nor_lower_wall_prop[i]
  dfnor$nor_total_arena_prop[i]<-dfnor$nor_lower_middle_prop[i] + 
     dfnor$nor_lower_left_prop[i] +               
     dfnor$nor_upper_middle_prop[i] +              
     dfnor$nor_upper_right_prop[i] +              
     dfnor$nor_upper_left_prop[i] +                
     dfnor$nor_center_left_prop[i] +              
     dfnor$nor_center_right_prop[i] +
     dfnor$nor_lower_right_prop[i]}
condition <- rep(c("wall" , "arena" , "center"))
value <- c(mean(dfnor$nor_total_wall_prop),mean(dfnor$nor_total_arena_prop),mean(dfnor$nor_center_middle_prop))
data <- data.frame(condition,value)

ggplot(data, aes(fill=condition, y=value, x = 1)) + 
    geom_bar(position="stack", stat="identity") +
    theme_classic()+
    scale_fill_manual(values = c("slateblue4", "mediumpurple3", "thistle3"))+
    ggtitle("NOVEL OBJECT Zones 
            (averaged across all)")


############SSB

cldata(pathtofile = "PLEASE INSERT PATH TO RAW SCORE SHEETS HERE /ssb", outputdataname = "dfssb", outputzonename = "ssb_list_of_zones_original", factor = "FALSE")
colnames(dfssb) <- paste("ssb", colnames(dfssb), sep = "_")

i<-1
for (i in 1:nrow(dfssb)){
  dfssb$ssb_total_investigation_prop[i]<-dfssb$ssb_positive_investigation_prop[i] + dfssb$ssb_aggressive_investigation_prop[i] + dfssb$ssb_rearing_prop[i]}
i<-1
for (i in 1:nrow(dfssb)){
  dfssb$ssb_total_prosocial_prop[i]<-dfssb$ssb_positive_side_by_side_contact_prop[i] + dfssb$ssb_huddling_prop[i] +  dfssb$ssb_allogrooming_prop[i]}
i<-1
for (i in 1:nrow(dfssb)){
  dfssb$ssb_total_aggression_prop[i]<- dfssb$ssb_biting_prop[i] + dfssb$ssb_chasing_prop[i]+ dfssb$ssb_being_chased_prop[i] + dfssb$ssb_pinning_prop[i]}

condition <- rep(c("neutral" , "aggression" , "prosocial", "interaction"))
value <- c(mean(dfssb$ssb_neutral_prop),mean(dfssb$ssb_total_aggression_prop),mean(dfssb$ssb_total_prosocial_prop), mean(dfssb$ssb_total_investigation_prop))
data <- data.frame(condition,value)

ggplot(data, aes(fill=condition, y=value, x = 1)) + 
    geom_bar(position="stack", stat="identity") +
    theme_classic()+
    scale_fill_manual(values = c("midnightblue", "royalblue3", "cornflowerblue", "lightskyblue1"))+
    ggtitle("SSB Zones 
            (averaged across all)")


############OSB

cldata(pathtofile = "PLEASE INSERT PATH TO RAW SCORE SHEETS HERE /osb", outputdataname = "dfosb", outputzonename = "osb_list_of_zones_original", factor = "FALSE")
colnames(dfosb) <- paste("osb", colnames(dfosb), sep = "_")

i<-1
for (i in 1:nrow(dfosb)){
  dfosb$osb_total_investigation_prop[i]<-dfosb$osb_positive_investigation_prop[i] + dfosb$osb_aggressive_investigation_prop[i] + dfosb$osb_rearing_prop[i]}
i<-1
for (i in 1:nrow(dfosb)){
  dfosb$osb_total_prosocial_prop[i]<-dfosb$osb_positive_side_by_side_contact_prop[i] + dfosb$osb_huddling_prop[i] +  dfosb$osb_allogrooming_prop[i]}
i<-1
for (i in 1:nrow(dfosb)){
  dfosb$osb_total_aggression_prop[i]<- dfosb$osb_biting_prop[i] + dfosb$osb_chasing_prop[i]+ dfosb$osb_being_chased_prop[i] + dfosb$osb_pinning_prop[i]}

condition <- rep(c("neutral" , "aggression" , "prosocial", "interaction"))
value <- c(mean(dfosb$osb_neutral_prop),mean(dfosb$osb_total_aggression_prop),mean(dfosb$osb_total_prosocial_prop), mean(dfosb$osb_total_investigation_prop))
data <- data.frame(condition,value)

ggplot(data, aes(fill=condition, y=value, x = 1)) + 
    geom_bar(position="stack", stat="identity") +
    theme_classic()+
    scale_fill_manual(values = c("darkgreen", "forestgreen", "darkolivegreen3", "honeydew2"))+
    ggtitle("osb Zones 
            (averaged across all)")


```

#generating homecage graphs (stats from SPSS)
```{r}

#COMPARISON OF TREATMENT

dfhome<-read_csv("PLEASE INSERT PATH TO RAW HOMECAGE DATA HERE")

dfhome$littertreatment<-paste(dfhome$litter, dfhome$treatment, sep = "_")
hometemp<-aggregate(cbind(parent_nursing, pup_self_grooming, pup_barrier_nose_poke, pup_barrier_traverse, parent_barrier_nose_poke, parent_barrier_traverse)~littertreatment,dfhome,mean)

for (i in 1:nrow(hometemp)){
hometemp$treatment[i]<-strsplit(toString(hometemp$littertreatment[i]), split = "_")[[1]][2]}

homelist<-c("parent_nursing", "pup_self_grooming", "pup_barrier_nose_poke", "pup_barrier_traverse", "parent_barrier_nose_poke", "parent_barrier_traverse")

i<-1
for (i in 1:length(homelist)){
  dftemp<-hometemp %>% select(treatment, homelist[i])
  dftemp$treatment<-as.factor(dftemp$treatment)
  colnames(dftemp)<-c('treatment', 'measurement')
  shapiro<-shapiro.test(dftemp$measurement)

p<- ggplot(dftemp, aes(x=treatment, y=measurement)) +
   geom_boxplot(lwd=1, fill = c("gray75", "gray95"))+
   theme_classic()+
   labs(x= "treatment",y=homelist[i]) + theme(axis.text.y = element_text(size = 14)) + theme(axis.text.x = element_text(size = 14)) + ggtitle(homelist[i])+stat_n_text()
 plot(p)}

```





